Hot and fresh: evidence of climate-related suboptimal water conditions for seagrass in a large Gulf coast estuary

### 2024-06-03 revision2/review_response.qmd

---
format: 
  docx:
    reference-doc: response_styles.docx
bibliography: ../refs.bib
csl: ../estuaries-and-coasts.csl

execute: 
  echo: false
  warning: false
  message: false
---

```{r}
#| eval: false
spelling::spell_check_files(here::here('revision2/review_response.qmd'))
```

@@

### AE comments:

Thank you for your revisions on this manuscript, I especially appreciate your well thought out responses to reviewer comments and also the changes made to the analyses (particularly the new GAM models and inclusion of the light attenuation data).  It will be a very nice contribution.  I have only a few minor comments that should be addressed:

*Response: Thank you once again for providing useful comments on our manuscript.  We have addressed them below and provide additional interpretation regarding the interaction terms in the GAMs.*

Use of p-values:  I do appreciate what ESCO is trying to address with the move away from our reliance on statistical significance and p-values.  However, I think in this case, it is useful to include both.  For your GAM models, you indicate some smoother relationships are weak or strong, and this is based on the p-value.  Here the p-values indicate the difference of the smoother from a constant straight line and it is an important way to evaluate their significance.  I suggest inserting the smoother p-values into the text in addition to the text that is already present (I found them in the tables but it would be easier in the text).  Further, I feel p-values are important for regression results, because you can have a significant regression (say p=0.01) but a low R2 value, indicating a relationship between the response and predictor is identified but there is lots of spread in the data.  So, I would also suggest inserting these. 
*Response: We have added p-values to the text when describing the relationship of predictors to the seagrass response variables.  Please see section 3.3 for updates.*

@@

For the GAM analysis, you may want to consider removing the interactions with year, to provide more power to define the GAM smoothers.  Most of these interactions are non-significant, and you’ve examined them individually, so you would be justified in removing these. You could mention you did these in preliminary models and no strong patterns were observed.  Or, if there is no difference you could just leave them.  Regardless, it is likely worthwhile to investigate.

*Response: We have removed all interaction terms from the GAMs and updated the figures, tables, and text.  The results were very similar and we agree it was not worth including these in the models.  We have made a note in the text that initial models evaluated interaction terms, but the results were similar.*

*Lines 370-373: Initial models also included a tensor product interaction term with a cubic regression spline that evaluated the potential interacting effects of each predictor with year. However, most of the interaction terms were not significant and were excluded from the final models.*

@@

Line 491: Change “model for the EPC “ to “seagrass cover…”

*Response: Changed*

Line 493: seagrass cover was associated with decreased light attenuation for both HB and MTB segments.  In HB, it was a fairly random pattern with not much biological meaning.  For MT though seagrass cover increased with increased light attenuation, but as you mention, this is still below the threshold.  Some clarification here might be useful.

*Response: The following was added on lines 489-493: "Associations with light attenuation were also observed for the HB (p < 0.001) and MTB (p = 0.004) bay segments, although the relationship in HB was flat beyond 0.75 m$^{-1}$ and opposite as expected for MTB (Figure 7, Table S7). As noted above, light attenuation in HB and MTB in recent years is well withing the range supportive of seagrass growth (Figure 2c), such that these associations may not be describing meaningful biological relationships."*

@@

Line 494: “associations were weak” – add smoother p-values here

*Response: Added.*

Line 498: both metric did not add additional value – however, it is significant for HB.  Not sure if this is relevant or not

*Response: Removing the interaction terms also showed a significant association in OTB, although again this is contrary to expectation as an increase in the metric was associated with an increase in seagrass.  The response to the following comment provides some explanation for this anomaly.*

@@

Line 555: “For the EPC model none of the stressor metrics were associated with seagrass change”.  This isn’t quite correct: – salinity OTB was, and also the general trends in increasing sg with increasing temperature should be acknowledged explicitly. Although these were not significant they do suggest a relationships.  I appreciate the distinction between the temperature metrics used for the different datasets as you outline, but what would the mechanism be?  Could the increase with temperature in the EPC data indicate increased vegetative growth in anomalous temperature conditions?  Several studies have observed this (DuBois et al. 2020, Wong and Dowd 2023 etc)

*Response: Yes, this was an explanation that was initially considered for the temperature metric, but not stated in original the text. It is also difficult to describe the positive association of seagrass change with the salinity metric in OTB.  Our description later in the paragraph regarding the spatial mismatch between the water quality and seagrass data is a possible explanation.*

*The following text was added to the discussion (lines 557-561): "The positive association of seagrass change with the temperature (or 'both') metric could be a signal of increased vegetative growth with temperature anomalies (e.g., Wong and Dowd 2023), whereas the positive association with the salinity metric in OTB is not easily explained since extreme precipitation events have been linked to seagrass loss in Tampa Bay (Greening and Janicki 2006)."*

@@

Line 611: I think this explanation is really important, can it be highlighted above too.  What is the depth difference and does this ameliorate solar heating effects experienced by the seagrass.

*Response: The mean depth of all EPC sampling stations (excluding LTB) is 3.88 meters, whereas the mean depth of points along each transect and sample date is 0.76 meters. The following was added to lines 363-365 in section 2.7 of the methods: "...; note that the seagrass transect data were collected at different locations and depths ($\bar{Z}$ = 0.76 m) than the EPC water quality data ($\bar{Z}$ = 3.88 m, excludes LTB not evaluated)."*

### 2024-06-03 cover-letter.qmd
revision2 done

Enclosed please find our revised manuscript, titled \`Hot and fresh: evidence of climate-related suboptimal water conditions for seagrass in a large Gulf coast estuary', to be considered as an original article in Estuaries and Coasts.

We sincerely appreciate the constructive comments provided by the editors and reviewers and have made every attempt to address these comments in our revision. The largest change to the manuscript was a reanalysis of the potential seagrass response to climate stressors, which included more flexible Generalized Additive Models and inclusion of additional predictors (i.e., time as a continuous variable, light attenuation).  Overall, our conclusions remain the same that climate stressors are potentially driving recent seagrass changes, although our more robust modeling approach provides greater weight of evidence for this conclusion.  Note that we have also made substantial revisions in response to comments from the co-editor in chief regarding statistical descriptions of results to focus more on effect sizes and magnitudes of change.

@@

Marcus W. Beck, Tampa Bay Estuary Program, St. Petersburg, Florida, USA, [mbeck@tbep.org](mailto:mbeck@tbep.org)

>>>>

Enclosed please find the second revision of our manuscript, titled \`Hot and fresh: evidence of climate-related suboptimal water conditions for seagrass in a large Gulf coast estuary', to be considered as an original article in Estuaries and Coasts.

We have addressed the remaining comments from Dr. Wong, as noted in our attached responses. In particular, we have removed the interaction terms from the GAMs and included appropriate statistics when describing the resulting relationships. Our main conclusions remain the same. Additional minor revisions are noted in our responses.

@@

Marcus W. Beck, Tampa Bay Estuary Program, St. Petersburg, Florida, USA, [mbeck\@tbep.org](mailto:mbeck@tbep.org)

### 2024-06-03 manu-draft.qmd
revision2 done

```{r}
#| results: hide
library(knitr)
library(here)
library(dplyr)
library(tibble)
library(tidyr)

@@

For comparison to seagrass, the annual metrics calculated from the EPC data were referenced to approximate periods of time between the annual seagrass transect surveys, as opposed to the calendar year for describing trends above. Bay segment stressor metrics were calculated as the average counts in each "transect year" from all stations in each segment from 1998 to 2022. Preliminary analyses evaluated different lagged associations between the stressor metrics and seagrass change, although initial results suggested no additional insight could be gained using lagged assessments compared to the transect year summaries. As such, the stressor metrics were compared to frequency occurrence (all species) each year by bay segment. GAMs were used to evaluate frequency occurrence in response to the independent variables, where the latter were the stressor metrics for temperature, salinity, or both. Additional predictors included year and light attenuation as estimated from Secchi depth [@tbep0696]. For 1998 to 2022, `r onb$epc` of the Secchi observations were recorded on the bottom. A single smooth term was used for each predictor using a thin plate regression spline, including a tensor product interaction term with a cubic regression spline that evaluated the potential interacting effects of each predictor with year. Two models were evaluated, one with the bottom temperature and salinity metrics together and another with the both metric. All models excluded Lower Tampa Bay because of minimal seagrass change over time. Additionally, total seagrass frequency occurrence was used as a response variable as an aggregate measure of community change to potential stressors given the likelihood that species-specific changes may be more difficult to model and that the majority of seagrass loss in recent years was dominated by *H. wrightii* in the upper bay segments (Figure [-@fig-seagrasschg]b). All GAMs were fit using restricted maximum likelihood evaluation [@wood11] with the *mgcv* R package \[@wood17.

Two other models were constructed to provide an additional weight-of-evidence for the FIM and PDEM temperature and salinity datasets relative to seagrass change. These models used direct measurements of bottom salinity and temperature as independent variables because the stressor metrics could not be calculated using the sampling designs from these monitoring programs (i.e., each sample was a distinct location). GAMs were used for the FIM data to evaluate annual average seagrass percent cover using year, temperature, and salinity as independent variables. Separate smoothed terms for each predictor and interactions with year for each predictor for each bay segment were included as above. GAMs for the PDEM data were constructed similarly, except only Old Tampa Bay was evaluated due to spatial limitations of the data. Neither the FIM nor PDEM models used light attenuation as a predictor variable given that a large percentage of Secchi observations were measured on the bottom (`r onb$fim` and `r onb$pinco`, respectively), providing further support that light environments have not been limiting for seagrasses. Model input data were further subset to include only months from July to October to describe seagrasses during the growing season and to reduce potential seasonal effects. Lastly, all data were averaged annually for the monthly subsets for comparability of sample size (i.e., power) with the EPC models. For the PDEM models, presence/absence was converted to frequency occurrence as the number of sites with seagrass in a year divided by the total number of sites and all data were subset to less than 2 meters to better characterize locations where seagrass occurs.

@@

```{r}
sgmodtxt <- modtxt_fun(sgmodsum)
```

GAMs to assess the potential effects of year, light attenuation, temperature and salinity on seagrass change provided some evidence that seagrass change was influenced by climate stressors, particularly for the FIM and PDEM datasets. The seagrass cover data (`r sgmodtxt$epcmod1`) had notable associations with year for the OTB and MTB bay segments and for light attenuation for the HB bay segment (@fig-sgmod1, Table S7). Associations of seagrass change with the temperature and salinity metrics were weak and somewhat contrary to expectation (i.e., an increasing partial effect on seagrass as the stressor metrics increased). The EPC model using the "both" stressor metric was similar (`r sgmodtxt$epcmod2`, Figure S7, Table S8), suggesting no added benefit of evaluating times when both temperature was above and salinity was below the threshold.

The GAMs for the FIM and PDEM datasets had notable associations with temperature and salinity stress, where an increase in temperature and a decrease in salinity was often associated with a decline in seagrass depending on bay segment. The FIM model (`r sgmodtxt$fimmod`, Figure [-@fig-sgmod2]a, Table S9) showed a slight decrease in seagrass with increasing temperature for HB and MTB, with a similar change as salinity decreased. The PDEM model (`r sgmodtxt$pincomod`, Figure [-@fig-sgmod2]b, Table S10) that evaluated only OTB showed a distinct decrease with seagrass as temperature increased, but no association with salinity. Some of the interaction terms of the predictors with year suggested the relationships with seagrass may have changed over time, but the results were not consistent between bay segments or models (Tables S7 to S10).

@@

Our relatively simple modeling approach provided some evidence that climate-related stressors impart some effect on recent seagrass losses in Tampa Bay. The models did not provide a consistent explanation that increasing temperature and decreasing salinity were key (or the sole) drivers. However, evaluating all models together as weight-of-evidence suggests there is value in considering multiple datasets and models to interpret noisy patterns and compounding ecological processes. Model results for the FIM and PDEM datasets both suggested that increasing temperature and decreasing salinity were associated with potential seagrass loss in recent years (@fig-sgmod2). For the EPC model, none of the stressor metrics were associated with seagrass change. An important distinction between the EPC and other models is that the former evaluated the number of days above/below thresholds each year to quantify annual temperature or salinity stress, whereas the latter evaluated observed temperature and salinity values at the time of seagrass sampling. Additionally, the EPC models evaluated water quality changes from data collected at relatively deeper locations than where seagrass typically grows (Figure [-@fig-map]b), suggesting a potential disconnect in relating the two. As such, both models attempted to describe the role of these stressors on potential seagrass change, but use different independent variables given the different sampling designs of each monitoring program. These differences highlight challenges describing autecological relationships in long-term datasets, while also demonstrating the utility of our weight-of-evidence approach to describe such relationships.

@@

![Partial effects of smoothers (rows) by bay segment (columns) from a Generalized Additive Model used to describe seagrass change relative to year, light attenuation, the number of days each year when bottom temperature was above 30 $^\circ$C, and the number of days each year when bottom salinity was below 25 ppt. The EPC data were used for the independent variables in the model. Partial effects describe the modeled association between each predictor and seagrass frequency occurrence after accounting for the effects of the other predictors. The model also included smoothed interaction terms with year (not shown). See Table S7 for additional model fit statistics. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay. `r modtxt_fun(sgmodsum)$epcmod1`](figs/sgmod1.png){#fig-sgmod1}

@@

![Partial effects of smoothers (rows) by bay segment (columns) from Generalized Additive Models evaluating seagrass changes versus year, bottom temperature, and bottom salinity for the (a) FIM (first three columns evaluating mean annual percent cover from 0-100, `r modtxt_fun(sgmodsum)$fimmod`) and (b) PDEM (right column evaluating annual frequency occurrence from 0-1 in OTB only, `r modtxt_fun(sgmodsum)$pincomod`) datasets. Partial effects describe the modeled association between each predictor and seagrass change after accounting for the effects of the other predictors. The models also included smoothed interaction terms with year (not shown). See Tables S9 and S10 for additional model fit statistics. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.](figs/sgmod2.png){#fig-sgmod2}

>>>>

```{r}
#| results: hide
library(knitr)
library(here)
library(dplyr)
library(tibble)
library(tidyr)
library(broom)
library(mgcv)

@@

```{r}
data('transect', package = 'tbeptools')
data('epcdata', package = 'tbeptools')
sgdep <- round(-1 * mean(transect$Depth, na.rm = T) / 100, 2)
epcdep <- epcdata %>% filter(bay_segment != 'LTB') %>% summarise(dep = mean(Total_Depth_m, na.rm = T)) %>% .[[1]] %>% round(2)
```

For comparison to seagrass, the annual metrics calculated from the EPC data were referenced to approximate periods of time between the annual seagrass transect surveys, as opposed to the calendar year for describing trends above. Bay segment stressor metrics were calculated as the average counts in each "transect year" from all stations in each segment from 1998 to 2022. Preliminary analyses evaluated different lagged associations between the stressor metrics and seagrass change, although initial results suggested no additional insight could be gained using lagged assessments compared to the transect year summaries. As such, the stressor metrics were compared to frequency occurrence (all species) each year by bay segment; note that the seagrass transect data were collected at different locations and depths ($\bar{Z}$ = `r sgdep` m) than the EPC water quality data ($\bar{Z}$ = `r epcdep` m, excludes LTB not evaluated). GAMs were used to evaluate frequency occurrence in response to the independent variables, where the latter were the stressor metrics for temperature, salinity, or both. Additional predictors included year and light attenuation as estimated from Secchi depth [@tbep0696]. For 1998 to 2022, `r onb$epc` of the Secchi observations were recorded on the bottom. A single smooth term was used for each predictor using a thin plate regression spline. Initial models also included a tensor product interaction term with a cubic regression spline that evaluated the potential interacting effects of each predictor with year. However, most of the interaction terms were not significant and were excluded from the final models. Two models were evaluated, one with the bottom temperature and salinity metrics together and another with the both metric. All models excluded Lower Tampa Bay because of minimal seagrass change over time. Additionally, total seagrass frequency occurrence was used as a response variable as an aggregate measure of community change to potential stressors given the likelihood that species-specific changes may be more difficult to model and that the majority of seagrass loss in recent years was dominated by *H. wrightii* in the upper bay segments (Figure [-@fig-seagrasschg]b). All GAMs were fit using restricted maximum likelihood evaluation [@wood11] with the *mgcv* R package \[@wood17.

Two other models were constructed to provide an additional weight-of-evidence for the FIM and PDEM temperature and salinity datasets relative to seagrass change. These models used direct measurements of bottom salinity and temperature as independent variables because the stressor metrics could not be calculated using the sampling designs from these monitoring programs (i.e., each sample was a distinct location). GAMs were used for the FIM data to evaluate annual average seagrass percent cover using year, temperature, and salinity as independent variables. Separate smoothed terms for each predictor and each bay segment were included as above. GAMs for the PDEM data were constructed similarly, except only Old Tampa Bay was evaluated due to spatial limitations of the data. Neither the FIM nor PDEM models used light attenuation as a predictor variable given that a large percentage of Secchi observations were measured on the bottom (`r onb$fim` and `r onb$pinco`, respectively), providing further support that light environments have not been limiting for seagrasses. Model input data were further subset to include only months from July to October to describe seagrasses during the growing season and to reduce potential seasonal effects. Lastly, all data were averaged annually for the monthly subsets for comparability of sample size (i.e., power) with the EPC models. For the PDEM models, presence/absence was converted to frequency occurrence as the number of sites with seagrass in a year divided by the total number of sites and all data were subset to less than 2 meters to better characterize locations where seagrass occurs.

@@

```{r}
sgmodtxt <- modtxt_fun(sgmodsum)
epc1pvl <- gam_pvals(sgmods$epcmod1)
epc2pvl <- gam_pvals(sgmods$epcmod2)
fimpvl <- gam_pvals(sgmods$fimmod)
pincopvl <- gam_pvals(sgmods$pincomod)
```

GAMs to assess the potential effects of year, light attenuation, temperature, and salinity on seagrass change provided some evidence that seagrass change was influenced by climate stressors, particularly for the FIM and PDEM datasets. The seagrass cover data (`r sgmodtxt$epcmod1`, EPC model) had notable associations with year for all bay segments (OTB: `r epc1pvl[['yr:bay_segmentOTB']]`, HB: `r epc1pvl[['yr:bay_segmentHB']]`, MTB: `r epc1pvl[['yr:bay_segmentMTB']]`) that followed long-term changes in @fig-seagrasschg. Associations with light attenuation were also observed for the HB (`r epc1pvl[['la:bay_segmentHB']]`) and MTB (`r epc1pvl[['la:bay_segmentMTB']]`) bay segments, although the relationship in HB was flat beyond 0.75 m$^{-1}$ and opposite as expected for MTB (@fig-sgmod1, Table S7). As noted above, light attenuation in HB and MTB in recent years is well within the range supportive of seagrass growth (Figure [-@fig-seagrasschg]c), such that these associations may not be describing meaningful biological relationships. Associations of seagrass change with the temperature (OTB: `r epc1pvl[['temp:bay_segmentOTB']]`, HB `r epc1pvl[['temp:bay_segmentHB']]`, MTB `r epc1pvl[['temp:bay_segmentMTB']]`) and salinity (OTB: `r epc1pvl[['sal:bay_segmentOTB']]`, HB `r epc1pvl[['sal:bay_segmentHB']]`, MTB `r epc1pvl[['sal:bay_segmentMTB']]`) metrics were not present or contrary to expectation (e.g., increase in seagrass with an increasing salinity metric in OTB, slight increase in seagrass cover with an increasing temperature metric in all bay segments). Similarly, the EPC model using the "both" stressor metric (`r sgmodtxt$epcmod2`, Figure S7, Table S8) had positive associations of the "both" metric with seagrass change in OTB (`r epc2pvl[['both:bay_segmentOTB']]`) and HB (`r epc2pvl[['both:bay_segmentHB']]`).

The GAMs for the FIM and PDEM datasets had notable associations with temperature and salinity stress, where an increase in temperature and a decrease in salinity was often associated with a decline in seagrass depending on bay segment. The FIM model (`r sgmodtxt$fimmod`, Figure [-@fig-sgmod2]a, Table S9) showed a slight decrease in seagrass with increasing temperature in MTB (`r fimpvl[['temp:bay_segmentMTB']]`) and a similar change as salinity decreased in MTB (`r fimpvl[['sal:bay_segmentMTB']]`) and OTB (`r fimpvl[['sal:bay_segmentOTB']]`). The PDEM model (`r sgmodtxt$pincomod`, Figure [-@fig-sgmod2]b, Table S10) that evaluated only OTB showed a slight decrease with seagrass as temperature increased (`r pincopvl[['temp']]`), but no association with salinity.

@@

Our relatively simple modeling approach provided some evidence that climate-related stressors impart some effect on recent seagrass losses in Tampa Bay. The models did not provide a consistent explanation that increasing temperature and decreasing salinity were key (or the sole) drivers. However, evaluating all models together as weight-of-evidence suggests there is value in considering multiple datasets and models to interpret noisy patterns and compounding ecological processes. Model results for the FIM and PDEM datasets both suggested that increasing temperature and decreasing salinity were associated with potential seagrass loss in recent years (@fig-sgmod2). For the EPC model, the salinity metric was positively associated with seagrass in OTB and all segments showed a slight increase in seagrass with increases in the temperature metric. The positive association of seagrass change with the temperature (or "both") metric could be a signal of increased vegetative growth with temperature anomalies [e.g., @wong2023], whereas the positive association with the salinity metric in OTB is not easily explained since extreme precipitation events have been linked to seagrass loss in Tampa Bay [@greening2006]. An important distinction between the EPC and other models is that the former evaluated the number of days above/below thresholds each year to quantify annual temperature or salinity stress, whereas the latter evaluated observed temperature and salinity values at the time of seagrass sampling. Additionally, the EPC models evaluated water quality changes from data collected at relatively deeper locations than where seagrass typically grows (Figure \[-@fig-map\]b), suggesting a potential disconnect in relating the two. As such, both models attempted to describe the role of these stressors on potential seagrass change, but use different independent variables given the different sampling designs of each monitoring program. These differences highlight challenges describing autecological relationships in long-term datasets, while also demonstrating the utility of our weight-of-evidence approach to describe such relationships.

@@

![Partial effects of smoothers (rows) by bay segment (columns) from a Generalized Additive Model used to describe seagrass change relative to year, light attenuation, the number of days each year when bottom temperature was above 30 $^\circ$C, and the number of days each year when bottom salinity was below 25 ppt. The EPC data were used for the independent variables in the model. Partial effects describe the modeled association between each predictor and seagrass frequency occurrence after accounting for the effects of the other predictors. See Table S7 for additional model fit statistics. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay. `r modtxt_fun(sgmodsum)$epcmod1`](figs/sgmod1.png){#fig-sgmod1}

@@

![Partial effects of smoothers (rows) by bay segment (columns) from Generalized Additive Models evaluating seagrass changes versus year, bottom temperature, and bottom salinity for the (a) FIM (first three columns evaluating mean annual percent cover from 0-100, `r modtxt_fun(sgmodsum)$fimmod`) and (b) PDEM (right column evaluating annual frequency occurrence from 0-1 in OTB only, `r modtxt_fun(sgmodsum)$pincomod`) datasets. Partial effects describe the modeled association between each predictor and seagrass change after accounting for the effects of the other predictors. See Tables S9 and S10 for additional model fit statistics. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay.](figs/sgmod2.png){#fig-sgmod2}
